---
title: "Chapter 2 - Installing packages"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE) # Don't eval any chunks here
```

> In this chapter we'll install Motus R packages as well as other packages we'll be using throughout these chapters.

## Installing Motus R packages
Two R packages have been developed for Motus users:

1. `motus`: provides functions for downloading and updating detection and deployment data, as well as for creating summary plots, and transforming (e.g. adding sunrise/sunset times) and analyzing Motus data.

2. `motusData`: provides sample datasets used in some of these articles

Motus users can install the latest stable versions of the R packages using `install.packages()` (see below). 
As with all R packages, you only need to install the packages once; 
after installation, you need to load each package (using `library()`) each time you open a new R session. 

Please note that some functionalities of the `remotes` package may require updated versions of R and RStudio. 
To avoid errors, please ensure you are using the most recent releases of [R](https://www.r-project.org/) **and** [RStudio](https://www.rstudio.com/products/RStudio/).
Update your R packages using `update.packages()` in the R console.

Begin by installing the required packages, if not already installed.

If you have used the older version of `motus` which included use of the `motusClient` package, it is recommended to first uninstall both packages.

```{r}
remove.packages(c("motus", "motusClient"))
```

Then proceed with the installation of the `motus` package

```{r}
install.packages("remotes")
library(remotes)

# install motus
install_github("MotusWTS/motus")

# install motusData package which contains sample datasets, e.g., vanishBearing
# used in Chapter 7
install_github("MotusWTS/motusData")

# Load the packages for use
library(motus)
library(motusData)
```

If you need to update the existing `motus` package, the process is the same:
```{r}
install_github("MotusWTS/motus")
```

If you want to know what version of the motus package you currently have installed:
```{r}
packageVersion("motus")
```


> If you are running into difficulties installing the `motus` package, please refer to the [`troubleshooting`](troubleshooting.html) section for tips.


## Installing other packages
Throughout these articles and examples, we use the [`tidyverse`](https://www.tidyverse.org/) collection of R packages for data science, including `tidyr`, `dplyr`, `ggplot2`, and `lubridate` (for managing and manipulating dates).
See the [`tidyverse` website](https://www.tidyverse.org/) for more information, or browse (or better still, thoroughly read) [R for Data Science](http://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham.
For mapping we also use the `rworldmap`, and `ggmap` packages.
These can be installed from CRAN, as follows:

```{r}
install.packages("maps")
library(maps)

install.packages("tidyverse")
library(tidyverse)

install.packages("rworldmap")
library(rworldmap)

install.packages("ggmap")
library(ggmap)
```

```{r}
library(motus)
library(maps)
library(tidyverse)
library(rworldmap)
library(ggmap)
```
Install but do not load the `plyr` package; 
you can use it directly for the handy `round_any()` function (with the code `plyr::round_any()`), but loading it can cause problems with `dplyr` functions:

```{r install plyr, eval = FALSE}
install.packages("plyr")
```

## Internal data processing

As an animal moves within the detection range of a Motus station, radio transmissions, or 'bursts', are detected by antenna(s) and recorded by a receiver. 
These raw detection data are either uploaded to the Motus database instantaneously via internet connection, or downloaded periodically from the receiver and uploaded to Motus manually.
Behind the scenes, various functions read and process the raw detections data to produce the tag detections file that users access using the R package (see [Chapter 3 - Accessing Data](03-accessing-data.html)). 
While most users will not need to call on the internal data processing functions, a complete list of functions within the Motus server R package can be found on the GitHub [motusServer repository](https://github.com/MotusWTS/motusServer). 

> **Next** [Chapter 3 - Accessing detections data](03-accessing-data.html)
> <span style = "font-size: 70%">(<a href = "index.html">Explore all articles</a>)</span>