---
title: "Chapter 2 - Installing packages"
author: Tara L. Crewe, Zoe Crysler, and Philip Taylor. Revisions by Steffi LaZerte and Denis Lepage
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE) # Don't eval any chunks here
```


## Installing Motus R packages
Two R packages have been developed for Motus users:

1. `motus`: provides functions for downloading and updated detections and deployment data, as well as for creating summary plots, and transforming (add sun rise/sun set times) and analyzing Motus data.

2. `motusData`: provides sample datasets used in some of these articles

Motus **users** can install the latest stable versions of the R packages using `install.packages()` (see below). 
As with all R packages, you only need to install the packages once; 
after installation, you need to load each package (using `library()`) each time you open a new R session. 

Please note that some functionalities of the `remotes` package may require updated versions of R and RStudio. 
To avoid errors, please ensure you are using the most recent releases of [R](https://www.r-project.org/) **and** [RStudio](https://www.rstudio.com/products/RStudio/).
Update your R packages using `update.packages()` in the R console.

Begin by installing the required packages, if not already installed.

If you have used the older version of `motus` which included use of the `motusClient` package, it is recommended to first uninstall both packages.

```{r}
remove.packages(c("motus", "motusClient"))
```

Then proceed with the installation of the `motus` package

```{r}
install.packages("remotes")
library(remotes)

# install motus
install_github("MotusWTS/motus")

# install motusData package which contains sample datasets, e.g., vanishBearing
# used in Chapter 7
install_github("MotusWTS/motusData")

# Load the packages for use
library(motus)
library(motusData)
```

If you need to update the existing `motus` package, the process is the same:
```{r}
install_github("MotusWTS/motus")
```

If you want to know what version of the motus package you currently have installed:
```{r}
packageVersion("motus")
```

### Troubleshooting the installation

Occasionally users run into problems while trying to install or update `motus`. Often this is related to problems with different versions of package dependencies. Here we suggest several solutions.

1. Update all packages during the installation

```{r}
library(remotes)
install_github("MotusWTS/motus", upgrade = "always")
```

2. If the installation of Motus generates errors saying that some of the existing packages cannot be removed, you can try to quit any R session, manually delete the problematic package folder from your R libraries and manually install the package again before trying to install `motus.` 
You can also try to set up a custom R library folder with `.libPaths()` and ensure that you have full write permissions on that folder, or try to start R in administrator (Windows) or SUDO mode (Linux/Ubuntu) and try installing again.

To set a custom library folder for installing new packages:
```{r}
.libPaths("C:/r-libraries/")
```

3. In some cases, it is easier to upgrade R itself by reinstalling the newest version of R: <https://cran.r-project.org/>. **Note:** While this results in a nice clean installation with fewer problems, it necessitates the re-installation of R packages which can be time-consuming.

4. If reinstalling R is not an option, you get an error related to packages built under a current version of R, AND updating your packages doesn't help, you can consider overriding the error with the following code. **Note:** This might help you install `motus` but may result in other problems. If possible, it's best to resolve the errors rather than ignoring them.

```{r}
library(remotes)
Sys.setenv("R_REMOTES_NO_ERRORS_FROM_WARNINGS"=TRUE)
install_github("MotusWTS/motus", upgrade = "always")
```


## Installing other packages
Throughout these articles and examples, we use the [`tidyverse`](https://www.tidyverse.org/) collection of R packages for data science, including `tidyr`, `dplyr`, `ggplot2`, and `lubridate` (for managing and manipulating dates).
See the [`tidyverse` website](https://www.tidyverse.org/) for more information, or browse (or better still, thoroughly read) [R for Data Science](http://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham.
For mapping we also use the `rworldmap`, and `ggmap` packages.
These can be installed from CRAN, as follows:

```{r}
install.packages("maps")
library(maps)

install.packages("tidyverse")
library(tidyverse)

install.packages("rworldmap")
library(rworldmap)

install.packages("ggmap")
library(ggmap)
```

```{r}
library(motus)
library(maps)
library(tidyverse)
library(rworldmap)
library(ggmap)
```
We also install but do not load the `plyr` package; 
we use it directly for the handy `round_any()` function (with the code `plyr::round_any()`), but loading it can cause problems with the `dplyr` functions:

```{r install plyr, eval = FALSE}
install.packages("plyr")
```

## Internal data processing

As an animal moves within the detection range of a Motus station, radio transmissions, or 'bursts', are detected by antenna(s) and recorded by a receiver. 
These raw detection data are either uploaded to the Motus database instantaneously via internet connection, or downloaded from the receiver and uploaded to Motus manually.
Behind the scenes, various functions read and process the raw detections data to produce the tag detections file that users access using the R package (see [Chapter 3 - Accessing Data](03-accessing-data.html)). 
While most users will not need to call on the internal data processing functions, a complete list of functions within the Motus server R package can be found on the GitHub [motusServer repository](https://github.com/MotusWTS/motusServer). 