---
title: "Database Variables"
---

```{r, include = FALSE}
motus:::get_sample_data()
options(width = 90)

library(motus)
library(tidyverse)
library(lubridate)
library(kableExtra)
library(glue)

tags <- tagme(176, update = TRUE, new = FALSE, dir = "./data")
```

## Variables

This table lists all variables present in .motus SQLite databases. 

**Notes:**

- `Variable Name` refers to the name of a variable (column/field) in a table
- `Table Name` reflects the tables (actual database tables) where that variable can be found
- `View`s refer to the database 'views' where that variable can be found 
  (views do not contain data themselves, but show data collected and arranged 
  from various tables)
- `Variable Name in View` reflects the name of the variable in a view. 
  Variables often are given new names in views
  - e.g., `alt` is a variable in the `gps` table, 
  but when put into the `alltagsGPS` view, it is called `gpsAlt`
- `Creation Comments` are specific comments about how a particular variable was
  created. This applies mostly to variables which are created on the fly, or, 
  might come for different tables. 
  
> Use the Search bar to narrow results down to a particular variable (e.g., `latitude`),
> table (e.g., `recvDeps`), or view (e.g., `alltags`)


```{r, echo = FALSE}
t <- tibble(table = DBI::dbListTables(tags)) %>%
  mutate(name_table = map(table, ~DBI::dbListFields(tags, .))) %>%
  unnest(name_table)

v <- system.file("extdata/fields_glossary.csv", package = "motus") %>%
  read.csv(skip = 1) %>%
  bind_rows(t) %>%
  arrange(name_table, view) %>%
  select(name_table, table, view, name_view, comment, description) %>%
  mutate(name_view = replace_na(name_view, "--"),
         view = replace_na(view, "--"),
         comment = replace_na(comment, "")) %>%
  group_by(name_table) %>%
  summarize(table = paste(unique(table), collapse = ", "),
            view = paste(unique(view), collapse = ", "),
            name_view = paste(unique(name_view), collapse = ", "), 
            comment = paste(unique(comment), collapse = ", ")) %>%
  mutate(name_view = str_remove_all(name_view, "(, --)|(--, )"),
         view = str_remove_all(view, "(, --)|(--, )"),
         comment = str_remove_all(comment, ", $"),
         name_table = replace_na(name_table, "--"),
         table = replace_na(table, "--"))

v %>%
  rename("Variable name" = "name_table", 
         "Table name" = "table",
         "Variable name in view" = "name_view",
         "Creation comments" = "comment") %>%
  rename_all(tools::toTitleCase) %>%
  DT::datatable(options = list(pageLength = nrow(v), dom = "ft"), 
                rownames = FALSE)
```

> **What Next?** [Explore all articles](index.html)
