---
title: "Removing deprecated batches"
---

```{r, include = FALSE}
library(motus)
library(tidyverse)

motus:::get_sample_data()
options(width = 90)
knitr::opts_chunk$set(cache = TRUE)
sql_motus <- tagme(176, update = TRUE, new = FALSE, dir = "./data/")
```

```{r, message = FALSE, eval = FALSE}
library(motus)
library(tidyverse)

sql_motus <- tagme(176, update = TRUE, new = TRUE, dir = "./data/")
```

As work progresses and data are fine-tuned, batches of data may be re-run on the 
Motus servers, and as a result, be assigned new batch numbers 
(see the section on [Reprocessing Data](motus-data.html#reprocessing-data) in the [Understanding Motus Data](motus-data.html) article).

This results is a disconnect between the user's local data and the data on the server.
When the user updates their data, they'll have the new batches, but will also have the older, *deprecated* batches.

Users can see which batches have been deprecated in the `deprecated` table:

```{r}
tbl(sql_motus, "deprecated")
```

Each `batchID` listed represented a deprecated batch. The column `removed` indicates
whether this batch has been removed from your data (`1`) or not (`0`). 

This table is updated every time you update your data with a call to `tagme()`.
If you want to update it again (without removing anything), you can use

```{r, collapse = TRUE}
sql_motus <- deprecateBatches(sql_motus, fetchOnly = TRUE)
```


To see where these batches are in your data, you can filter for the IDs in a 
specific table

```{r}
tbl(sql_motus, "alltags") %>%
  filter(batchID %in% c(6000, 6001, 6002))
```

Deprecated batches can also be removed with the `deprecateBatches()` function,
which, by default, fetches the update-to-date record of deprecated batches and then removes them.

However, **once removed, deprecated batches are gone for good**.
It is advisable to backup your dataset before proceeding 
([see below](#why-not-remove-deprecated)).

Just in case, `deprecateBatches()` will ask you if you are sure you want to 
remove the batches before proceeding. 

```{r, eval = FALSE}
sql_motus <- deprecateBatches(sql_motus)
```

```
You are about to permanently delete up to 6 deprecated batches from project-176.motus
Continue? 

1: Yes
2: No

Selection: 1
```

```{r, echo = FALSE, collapse = TRUE}
sql_motus <- deprecateBatches(sql_motus, ask = FALSE)
```


After removal, you can see that the batches have been removed from the data

```{r}
tbl(sql_motus, "alltags") %>%
  filter(batchID %in% c(6000, 6001, 6002))
```

Also that the `deprecated` table now lists `removed` as 1.

```{r}
tbl(sql_motus, "deprecated")
```


## Why **not** remove deprecated batches? {#why-not-remove-deprecated}
In active projects it is a good idea to remove deprecated batches routinely,
to ensure your data is update to date and as accurate as possible.

However, once removed, deprecated batches are gone for good. 
If you have an analysis or publication based on older data versions, 
this analysis is no longer repeatable as you no longer have the original batches. 

It is therefore advisable to keep a copy of the database (i.e. the `XXXX.motus`
file) you need for a specific analysis backed up and static 
(i.e. no updates and no removal of deprecated batches).


> **What Next?** [Explore all articles](index.html)
