---
title: "Working with GPS points"
---

```{r, include = FALSE}
motus:::local_auth()
if(!dir.exists("data")) dir.create("data")
file.copy(system.file("extdata", "gps_sample.motus", package = "motus"),
          "./data/")
gps_tags <- DBI::dbConnect(RSQLite::SQLite(), "./data/gps_sample.motus")

# Get and update sample data to avoid messages later
sql.motus <- motus:::get_sample_data()
```

Some tags also collect GPS points, which can be used for more precise locations (as opposed to the receiver location `recDeployLat`/`recDeployLon`). 
In this article we will show you how to efficiently work with these GPS points. 


First, let's get set up. 
The motus.sample data (project 176) doesn't have GPS points, so in these next examples we'll be using a different data base called `gps_tags`. 

```{r, message = FALSE}
library(motus)
library(tidyverse)
library(lubridate)

gps_tags
```

If your data has gps point, you should be able to follow along using your own data (replace `PROJECT_NUMBER` with the number corresponding to your project.
```{r, eval = FALSE}
gps_tags <- tagme(PROJECT_NUMBER, update = TRUE, new = TRUE)
```

GPS points are stored in the `gps` table. 

```{r}
tbl(gps_tags, "gps")
```

## Working with GPS points

You can work with GPS points in one of two ways:

1. For small data bases - Use the `alltagsGPS` view directly (can be slow)
2. For large data bases - Filter first, then use the `addGPS()` function to match GPS points to your hits


### `alltagsGPS` view

The `alltagsGPS` view combines hits and GPS points
```{r}
tbl(gps_tags, "alltagsGPS") %>%
  select(hitID, runID, batchID, ts, gpsLat, gpsLon, gpsAlt)
```

Note that not all hits have a GPS point associated, but we can filter to those that do:

```{r}
tbl(gps_tags, "alltagsGPS") %>%
  select(hitID, runID, batchID, ts, gpsLat, gpsLon, gpsAlt) %>%
  filter(!is.na(gpsLat))
```

> Note: The `alltagsGPS` view is the same as the `alltags` view but includes GPS points. 
> Because of this, the `alltagsGPS` view can be slower to work with, particularly if you have a large database.


### Filtering then adding GPS data


For example, let's work with a subset of the alltags view, including only King Rails.

```{r}
rails <- tbl(gps_tags, "alltags") %>%
  filter(speciesEN == "King Rail")
```

Now let's retrieve the daily median location of GPS points for these data.
Note that we use both the original database `gps_tags` as well as the data subset `rails`. 
Also note that the `getGPS()` function requires the original, numeric `ts` column, so if you want a date/time column it's best to rename it (i.e. `time = as_datetime(ts)`).

```{r}
index_GPS <- getGPS(src = gps_tags, data = rails)
```

This table is an index matching GPS points to specific `hitID`, so the next step is to join it into your data subset using the `left_join()` function from the `dplyr` package.
Note that at this point, we need to use `collect()` to ensure that `rails` is 'flat' (i.e. a data frame, not a database; see [Converting to flat data](03-accessing-data.html#converting-to-flat-data) for more details).

```{r}
rails_GPS <- left_join(collect(rails), index_GPS, by = "hitID")
```

We can subset the columns to see if it worked as expected 
(we filter to non-missing, because not all hits have a GPS point when using the
default matching).

```{r}
rails_GPS %>%
  select(hitID, runID, batchID, ts, contains("gps")) %>%
  filter(!is.na(gpsLat))
```



#### More ways of matching GPS points

By default, `getGPS()` matches GPS points to hits by date. 
However, we can match GPS locations to `hitID`s according to one of several different time values, specified by the `by` argument.

`by` can be one of three options:

1. the median location within **`by = X`** minutes of a `hitID`
    - here, `X` can be any number greater than zero and represents the size of the time block in minutes over which to calculate a median location
    - be aware that you should ideally not chose a period smaller than the frequency at which GPS fixes are recorded, or some hits will not be associated with GPS

For example, the median location within 60 minutes of a `hitID`.
```{r}
index_GPS <- getGPS(src = gps_tags, data = rails, by = 60)
index_GPS
```    
    
    
2. **`by = "daily"`** median location (**default**, used in first example)
    - similar to `by = X` except the duration is 24hr (same as `by = 1440`)
    - this method is most suitable for receiver deployments at fixed location.
    
```{r}
index_GPS <- getGPS(src = gps_tags, data = rails, by = "daily")
index_GPS
```
    
3. or the **`by = "closest"`** location in time
    - individual GPS lat/lons are returned, matching the closest `hitID` timestamp
    - this method is most accurate for mobile deployments, but is potentially slower than `by = X`.
    - you can also specify a `cutoff` which will only match GPS records which are within `cutoff = X` minutes of the hit. This way you can avoid having situations where the 'closest' GPS record is actually days away.

For example, the closest location in time noted within 2 hours of a hit.
```{r}
index_GPS <- getGPS(src = gps_tags, data = rails, by = "closest", cutoff = 120)
index_GPS
```

To keep all `hitID`s, regardless of whether they match to GPS data or not, use the argument `keepAll = TRUE`.
This results in `NA` for `gpsLat`, `gpsLon` and `gpsAlt` where there is no corresponding GPS hit (otherwise the hit is omitted). 

```{r}
index_GPS <- getGPS(src = gps_tags, data = rails, keepAll = TRUE)
index_GPS
```


## Using GPS locations

Now that we have our GPS data (either through loading `alltagsGPS` or using the `getGPS()` function), 
we can use these coordinates when [cleaning our data in Chapter 5](05-data-cleaning.html). 

In Chapter 5, we used the receiver deployment latitude and longitude: `recvDeployLat`, `recvDeployLon`.

However, now that you have gps data, you can create receiver latitude and longitude variables (`recvLat`, `recvLon`, `recvAlt`) based on the coordinates recorded by the receiver GPS (`gpsLat`, `gpsLon`, `gpsAlt`), and where those are not available, infilled with coordinates from the receiver deployment metadata (`recvDeployLat`, `recvDeployLon`, `recvDeployAlt`). 

Missing GPS coordinates may appear as `NA` if they are missing, or as `0` or `999` if there was a problem with the unit recording.

However, as we are changing values in the data, we'll need to [flatten the data first](03-accessing-data.html#converting-to-flat-data).

Using the example from Chapter 5 starting with [Checking receivers](05-data-cleaning.html#checking-receivers)
```{r}
sql.motus <- tagme(176, update = TRUE, dir = "./data/")

df.alltagsGPS <- tbl(sql.motus, "alltagsGPS") %>%
  mutate(recvLat = if_else(is.na(gpsLat) | gpsLat %in% c(0, 999), 
                           recvDeployLat, gpsLat),
         recvLon = if_else(is.na(gpsLon) | gpsLon %in% c(0, 999), 
                           recvDeployLon, gpsLon),
         recvAlt = if_else(is.na(gpsAlt), recvDeployAlt, gpsAlt)) %>%
  collect()  # Flatten the data

# Take a look
select(df.alltagsGPS, hitID, recvLat, recvLon)
```

[Continuing on in Chapter 5](05-data-cleaning.html#checking-receivers) with these values, you would then replace all instances of `recvDeployLat` with `recvLat` and `recevDeployLon` with `recvLon`.

You would also use this flattened data frame, `df.alltagsGPS`, rather than the un-flattened `tbl.alltagsGPS`.

```{r include = FALSE}
unlink("data/project-176.motus")
unlink("data/gps_sample.motus")
```

> **What Next?** [Explore all articles](index.html)
