---
title: "Get started"
---

<!-- Parts of this were originally at the end of Chapter 3 - Accessing Data -->

```{r setup, include = FALSE}
# Get and update sample data to avoid messages later
sql.motus <- motus:::get_sample_data()
```

This is a quick introduction to downloading Motus data. Definitely check out the various [articles](../index.html) for more details!

Let's get started by loading the packages we'll use (see also [Chapter 2 - Installing packages](02-installing-packages.html).)
```{r, message = FALSE}
library(motus)
library(dplyr)
library(lubridate)
```

Now we'll download some data. 
Here we're downloading project 176 (sample data, use user name and password 'motus.sample').
You can use your own project number or receiver name.

We use the `tagme()` function to download the data
```{r}
sql.motus <- tagme(176, new = FALSE, update = TRUE, dir = "./data/")
```

`tagme()` includes several options. Here we're using:

- `new = FALSE` means the data base already exists, if it doesn't, change `FALSE` to `TRUE`
- `dir = "./data/"` means the data base will be stored in the `data` folder inside your current working directory

`tagme()` stores a `project-176.motus` SQLite data base in `./data/`.

We can access parts of the data base by referencing the SQL object we created, `sql.motus`

Here, we'll retrieve the `alltags` view from the SQLite database using the `tbl()` function from the `dplyr` package
```{r}
tbl.alltags <- tbl(sql.motus, "alltags")
```

Next, we can convert this to a data frame (a flat file; see [Converting to flat data in Chapter 3](03-accessing-data.html#converting-to-flat-data))

Convert to flat (`collect()`) and create a date/time column `time` from the numeric timestamps, `ts`.
```{r}
df.alltags <- tbl.alltags %>%
  collect() %>%
  mutate(time = as_datetime(ts))
```
 
If you want to save this flat file, you can export as RDS (see [Exporting detections in Chapter 3](03-accessing-data.html#exporting-detections))
```{r}
saveRDS(df.alltags, "my_motus_data.rds")
```

Use caution when producing a flat file as using the full suite of fields can use a lot of memory.
This can slow R down considerably when dealing with large datasets. 

## Workflow summary
 
For your own data we suggest creating a script (or scripts) with the following workflow:

1. Download/update your data (see [Chapter 2 - Installing packages](02-installing-packages.html), [Chapter 3 - Accessing detections data](03-accessing-data.html))
2. Select variables of interest for the table you are working with (typically `alltags`) (see [Chapter 3 - Accessing detections data](03-accessing-data.html), [Chapter 4 - Tag and receiver deployments](04-deployments.html))
3. Initial cleaning (see [Chapter 4 - Tag and receiver deployments](04-deployments.html) and [Chapter 5 - Data cleaning](05-data-cleaning.html))
4. Output the resulting data as an .rds file (see [Exporting detections in Chapter 3](03-accessing-data.html#exporting-detections)). 
We suggest using RDS instead of CSV, because the RDS format preserves the underlying structure of the data (e.g. times stay as times). If you want to export your data to another program, then a CSV format might be preferred.  

```{r include = FALSE}
unlink("my_motus_data.rds")
```

> **What next?** Check out the walkthrough starting with [Chapter 1 - Introduction](01-introduction.html)
> <span style = "font-size: 70%">(<a href = "index.html">Explore all articles</a>)</span>

